Relationship Between Common Stream Health Indicators

### 2021-11-12 products/manuscript/Manuscript.Rmd
Project_p4 - CC

```{r, echo=FALSE, results='hide', message=FALSE}
#Load important packages:
library(tidyverse)
library(readxl)
library(here)
library(png)
library(knitr)
```

```{r, warning=FALSE, echo=FALSE, results='hide'}
#Lodaing data and providing a summary for the raw data
data_location <- here::here("data/raw_data/UOWN_data_master_spring2021 (2).xlsx")
rawdata <- readxl::read_excel(data_location)
glimpse(rawdata)
```

```{r, echo=FALSE, results='hide'}
#Set the location of the desired dataframes
MIDO_location <- here::here("data","processed_data","MIDO.RDS")
NORO_location <- here::here("data","processed_data","NORO.RDS")
BICO_location <- here::here("data","processed_data","BICO.RDS")

#Load in the Dataframes
MIDO <- readRDS(MIDO_location)
NORO <- readRDS(NORO_location)
BICO <- readRDS(BICO_location)
```

@@

### Below is the Linear Regression analysis from the Exploratory Data Analysis

#### MIDO

_Outcome: Biological Score_
```{r, echo=FALSE}
#Define data location
MIDO_bs_lm_location <- here::here("results","MIDO_bs_lm.png")
#Load in saved image from results folder
include_graphics(MIDO_bs_lm_location)
```
_Outcome: E. coli (cfu)_
```{r, echo=FALSE}
#Define data location
MIDO_ecfu_lm_location <- here::here("results","MIDO_ECFU_lm.png")
#Load in the data
include_graphics(MIDO_ecfu_lm_location)
```

#### BICO

_Outcome: Biological Score_
```{r, echo=FALSE}
#Define data location
BICO_bs_lm_location <- here::here("results","BICO_bs_lm.png")
#Load in saved image from results folder
include_graphics(BICO_bs_lm_location)
```
_Outcome: E. coli (cfu)_
```{r, echo=FALSE}
#Define data location
BICO_ecfu_lm_location <- here::here("results","BICO_ECFU_lm.png")
#Load in the data
include_graphics(BICO_ecfu_lm_location)
```

#### NORO

_Outcome: Biological Score_
```{r, echo=FALSE}
#Define data location
NORO_bs_lm_location <- here::here("results","NORO_bs_lm.png")
#Load in saved image from results folder
include_graphics(NORO_bs_lm_location)
```
_Outcome: E. coli (cfu)_
```{r, echo=FALSE}
#Define data location
NORO_ecfu_lm_location <- here::here("results","NORO_ECFU_lm.png")
#Load in the data
include_graphics(NORO_ecfu_lm_location)
```
</br>

@@

_Moving forward, I want to focus in on just the MIDO dataset. As you can tell, working with many variables in three different data sets can be quite cumbersome to read through. For ease of interpretation, the MIDO data set will serve as the data frame where I test how robust the linear fit models actually are for the data. MIDO is the only data set large enough that when we split the data set into a "test" and "train" data set, I actually have confidence in the resulting model fit analysis. For comparison, MIDO has an n = 88; NORO has an n = 35; and BICO has an n = 23. The only thing we are missing by dropping the other sites is assessing site-by-site comparison, which I am choosing not to focus on anyway._

@@

_The results indicated that only the relationship between E. Coli (cfu) and Conductivity were significant for both the train data (p = 1.374747e-05) and test data (p = 1.89214e-06). However, both models had high RMSE values, indicating that the models did not fit the data well and could not be used to predict E. coli (cfu) from conductivity, despite a significant relationship._

@@

## Summary and Interpretation

## Strengths and Limitations

@@

>>>>

@@

@@

_Following correlation analysis during exploration, focus was shifted to just the MIDO dataset. As you can tell, working with many variables in three different data sets can be quite cumbersome to read through. For ease of interpretation, the MIDO data set will serve as the data frame where I test how robust the linear fit models actually are for the data. MIDO is the only data set large enough that when we split the data set into a "test" and "train" data set, I actually have confidence in the resulting model fit analysis. For comparison, MIDO has an n = 88; NORO has an n = 35; and BICO has an n = 23. The only thing we are missing by dropping the other sites is assessing site-by-site comparison, which I am choosing not to focus on anyway._

@@

_Finally, LASSO analyses were performed with either biological score or E. coli cfu as the outcome of choice. The LASSO model analyzed which combination of stream health descriptive variables are the best for predicting either biological score or E. coli cfu values for the MIDO site through time._

_The results indicated that only the relationship between E. Coli (cfu) and Conductivity were significant for both the train data (p = 1.374747e-05) and test data (p = 1.89214e-06). However, both models had high RMSE values, indicating that the models did not fit the data well and could not be used to predict E. coli (cfu) from conductivity, despite a significant relationship. To test how well the LASSO models performed and how well they fit our data, Root Mean Squared Error (RMSE) was calculated for each model produced and for predicted values of MIDO stream health variables, using the LASSO models. Residuals between actual and predicted values were calculated to show how well the models could predict values. Finally, all RMSE values (actual and predicted data models) were compared to a null model for both biological score and E. coli cfu to evaluate model performance._

_The null linear models for biological score and E. coli cfu produced a RMSE = 6.63, with a standard deviation of 0.13 and RMSE = 0.52, with a standard deviation of 0.014, respectively. Results from the biological score LASSO machine learning model indicated that model 47 is the best performing models (RMSE = 6.50; STE = 0.16). However, based on comparative RMSE scores, model 47 does not really perform any better than the null model. This indicates that it is a bad fit to the MIDO data. Results from the E. coli cfu LASSO machine learning model indicated that model 41 is the best performing models (RMSE = 0.50; STE = 0.012). However, based on comparative RMSE scores, model 41 does not really perform any better than the null model. This indicates that it is a bad fit to the MIDO data._

@@

_Based on the linear modeling exploratory analysis and LASSO modeling for best combination of predictors, the results indicate that combinations of the variables biological score, E. coli cfu, nitrate concentration, conductivity, and turbidity are not able to predict values for biological score or E. coli cfu. This indicates that one cannot predict one measure of stream health from another. This is seen by a lack of consistent significant correlations between each stream health predictor and either biological score or E. coli cfu as well as LASSO modeling to find the ideal combination of stream health variables to predict biological score or E. coli cfu. Given these results, it is likely that the data was biased due to the  inconsistent and opportunistic sampling efforts done by citizen science [find citation for lack of reliability for citizen science data]. When examining the data set, many values were observed to be missing, with a disproportionate amount of extent observations in what I suspect to be the more easily collected stream health variables. For example, significantly more NA observations were included in the biological score column relative to variables like turbidty and conductivity. This is because both turbidity and conductivity can be measured in situ with a meter in a matter of minutes, while biological score is calculated following collection and identification of macroinvertebrate families from samples [find reference for biological score assessment methodology]. Therefore, biological score is much harder to measure. Additionally, E. coli cfu is processed by culturing E. coli from samples and generally requires a trained scientist to do [find reference for E. coli cfu qunatification methods]. While confidence in values may be greater in this case, there is know a shortage of labor due to need for a specialized skill set to process and quantify E. coli cfu values._

@@

_Based the results of this study, I would recommend that existing methodology for UOWN stream health characterization and monitoring be re-evaluated. The issue being that there seems to be a disparity in sampling effort, both in quality of samples and time spent sampling, for each different stream health metric. The result being that the term "healthy stream" does not have a unified definition in terms of the metrics used to quantify stream health. I would suggest future scientific studies be conducted to assess to relationship between stream health metrics in order to be able to predict metrics like biological score and E. coli cfu so citizen based monitoring does not have to collect these data while still having easier to measure variables that can be used to describe stream health in terms of biological score and E. coli cfu._
